---
title: "First Steps with the Phylogenetic Ornstein-Uhlenbeck Mixed Model"
author: "Venelin Mitov"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{First Steps with the Phylogenetic Ornstein-Uhlenbeck Mixed Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
references:
- id: mitov2016a
  title: The heritability of pahtogen traits - definitions and estimators
  author:
  - family: Mitov
    given: Venelin
  - family: Stadler
    given: Tanja
  container-title: BiorXiv
  URL: 'https://doi.org/10.1101/058503'
  DOI: 10.1101/058503
  type: article-journal
  issued:
    year: 2016
    month: 6
---

Here, we introduce the R-package **poumm** - an implementation of the Pylogenetic Ornstein-Uhlenbeck Mixed Model (POUMM) for univariate traits [@mitov2016a]. 

The POUMM is used to model the evolution of a continuous trait, $z$, along a phylogenetic tree, assuming that $z$ is the sum of a genetic (heritable) component, $g$, and a non-heritable (environmental) component, $e$: $z = g + e$. Within the family of the phylogenetic comparative methods, the , a long-term tendency towards a global optimum, i.e. stabilizing selection. As a member of phylogenetic comparative method, it represents a generalisation of the Phylogenetic Mixed Model (PMM) [@butler]

The **poumm** package comes with an efficient model likelihood calculation that scales to large phylogenies (currently tested with up to 100000 tips) and, if needed, can perform the likelihood calculation using the package **Rmpfr** for high floating point precision^[Note: The high-precision likelihood calculation is usually needed in case of accumulating numerical errors due to extreme values of some model parameters alpha, theta, sigma or sigmae or extreme (i.e. very long or very short) branch lengths in the phylogeny. As the Rmpfr package is currently not supported on all linux platforms, the poumm package doesn't explicitly depend on it, but you will need to install it manually and import it in case you specify usempfr=2 in the call to likVTreeOU.] 
Currently, the package depends on the package **ape** and the package **data.table**. The most important functions of the package are "likVTReeOU" (see ?likVTreeOU), which performs likelihood calculation, "mlOUTree" (type ?mlOUTree), which fits POUMM to a tree and data at the tips using maximum likelihood and mcmcOUTree, which runs an MCMC fit to a tree and data. Although the algorithm for likelihood calculation has been extensively tested and applied on various real as well as synthetic data, the package is still in a beta-version for the following reasons:

* The signature of the functions in the package may change;
* Some of the functions are not yet documented. 



# Installing the package
```{r install, eval=F}
install.packages('poumm_0.1.tgz', type='binary', repos=NULL)
```

# A Hello World example

```{r example, cache=TRUE}
library(poumm)

set.seed(1)
# Number of tips
N <- 1000
# true parameters
alpha <- 1
theta <- 4.5
sigma <- 2
sigmae <- 2

tree <- rtree(N)
tree$edge.length <- tree$edge.length/10

g <- generateTraitOU(tree, g0=6, alpha=alpha, theta=theta, sigma=sigma)
v <- rnorm(N, mean=0, sd=sigmae)+g[1:N]

#likelihood at the true value
lik <- likVTreeOU(v, tree, alpha, theta, sigma, sigmae, log=T, distgr='maxlik')

# setting the maxit parameter to a very small value (i.e. 2) while the 
# option 'hybrid' is on, would transform the pso algorithm into a gradient
# descent with a number of different original locations. 
fit <- mlOUTree(v=v, tree=tree, distgr='maxlik')

# maximum likelihood fit:
print(fit$par)

# ml likelihood value versus value at the original true params
print(c(-fit$value, lik))

# estimating broad-sense heritability
print(c(calcH2AtTime(fit$par[1], fit$par[3], fit$par[4], t=Inf, tm=0), 
        calcH2AtTime(alpha, sigma, sigmae, t=Inf, tm=0)))
```


# Further information and help
For further information, look in the package help-pages. If you cannot find the answer, please write me to venelin.mitov@bsse.ethz.ch. 




Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style

## Vignette Info

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))

# References
